# Cite This Data

This is the **Data Sources & Attribution** page — the one answer to "how do I cite CalCOFI
data?". Every CalCOFI product points here rather than restating it: the release itself has a
citation, and every dataset inside it carries its own — checked against the source's own
authority, never invented (the attribution contract lives in
[CalCOFI/workflows](https://github.com/CalCOFI/workflows/blob/main/CLAUDE.md)).

::: {.callout-important}
## Two things to cite, always

If your work touches the CalCOFI Integrated Database, cite **both**: the release
(the database as a whole — the assembly, the schema, the QA/QC) *and* every
individual dataset your analysis actually used (the people who collected and
curated the data). Citing only the release erases the contributing datasets;
citing only a dataset erases the integration work that made it queryable
alongside fifteen others.
:::

```{r}
#| label: setup
#| include: false
librarian::shelf(dplyr, gt, here, jsonlite, knitr, readr, quiet = TRUE)

# the promoted release, as snapshotted by libs/pre-render.R (the one place the
# network is touched at render time)
snap    <- jsonlite::fromJSON(here("data/_snapshot.json"))
catalog <- jsonlite::fromJSON(here("data/release/catalog.json"), simplifyVector = FALSE)
ds      <- readr::read_csv(here("data/release/dataset.csv"), show_col_types = FALSE)
```

## Citing the database release

```{r}
#| label: release-citation
#| echo: false
#| results: asis

# catalog.json carries the release's own citation (calcofi4db::release_citation()
# writes it at the freeze); cc_cite() in R and Python read the same field, so
# nothing here builds a citation string by hand.
cat(catalog$citation, "\n")
```

Every release also has a DOI of its own, minted on Zenodo when the release is tagged, and one
**concept DOI** that always resolves to the newest release. Cite the version DOI in a paper — it
names exactly the bytes you used — and the concept DOI where a living pointer is wanted. Both are
in `catalog.json` (`doi`, `concept_doi`) and in the release notes' *How to cite*:

```{r}
#| label: dois
#| echo: false
#| results: asis
cat(sprintf("- this release, `%s`: <https://doi.org/%s>\n- the concept DOI, every release: <https://doi.org/%s>\n",
            catalog$version, catalog$doi, catalog$concept_doi))
```

## Citing an individual dataset

Every dataset in the release carries its own `citation_main`, a registered `license`, and — where
the source gives one — a `doi`, an `acknowledgement` and a `contact`. The columns are `dataset.doi`,
`dataset.license_url`, `dataset.acknowledgement`, `dataset.contact` and `dataset.source_accessed`
(measured at release, never typed; see [the database](db.qmd)). They have shipped since
v2026.09.04. The table is the `dataset` table of the promoted release, `r snap$version`.

```{r}
#| label: dataset-table
#| echo: false

cols <- intersect(
  c("provider", "dataset", "dataset_name", "citation_main", "license",
    "license_url", "doi", "pi_names", "acknowledgement", "contact",
    "source_accessed", "link_calcofi_org", "link_data_source"),
  names(ds))

ds |>
  dplyr::select(dplyr::all_of(cols)) |>
  dplyr::arrange(provider, dataset) |>
  gt::gt() |>
  gt::cols_label(.list = list(
    provider = "provider", dataset = "dataset", dataset_name = "name",
    citation_main = "citation", license = "license", license_url = "license URL",
    doi = "DOI", pi_names = "PI(s)", acknowledgement = "acknowledgement",
    contact = "contact", source_accessed = "source accessed",
    link_calcofi_org = "calcofi.org", link_data_source = "data source")[cols]) |>
  gt::fmt_url(columns = dplyr::any_of(c("link_calcofi_org", "link_data_source")),
              label = "↗") |>
  gt::sub_missing(missing_text = "") |>
  gt::tab_options(table.width = gt::pct(100), table.font.size = "0.85em")
```

## Citing what you downloaded

Every figure, CSV or Parquet subset you export from a CalCOFI product should
carry citations for **only the datasets it actually used**, plus the release.
`cc_cite()` accepts a `dataset_key` vector or a query result data frame — it
resolves exactly the datasets present, in first-occurrence order, and errors
naming any `dataset_key` it does not recognize rather than silently dropping
one.

::: {.panel-tabset}

## R

```{r}
#| label: cite-r
#| eval: false
#| echo: true
library(calcofi4r)                # remotes::install_github("calcofi/calcofi4r")
con <- cc_get_db()                # the promoted release, over https

cc_cite(con = con)                # every dataset in the release, release citation first
cc_cite("calcofi_dic", con = con) # just the datasets a query touched

# from a query result directly (must carry a dataset_key column)
dplyr::tbl(con, "obs") |>
  dplyr::filter(dataset_key == "calcofi_dic") |>
  head(5) |>
  dplyr::collect() |>
  cc_cite(con = con)

# a .bib file for a paper's reference manager
cc_cite(format = "bibtex", con = con) |> cat()
```

## Python

```python
# pip install "calcofi4py @ git+https://github.com/CalCOFI/calcofi4py"
import calcofi4py as cc

con = cc.cc_get_db()

cc.cc_cite()                     # every dataset, release citation first
cc.cc_cite("calcofi_dic")        # just one dataset
cc.cc_cite(format="bibtex")      # a .bib file
```

:::

The [Explorer](https://calcofi.io/explore/) does the same from its *Cite this data* button — the
release citation plus one per dataset in the view, as text or BibTeX — and every CSV it exports
carries a `dataset_key` column for exactly this reason.

## Reaching a data provider

Each dataset's `contact` (a provider-chosen URL or `mailto:` link) is in the
table above. To share a derived product, ask a question about the underlying
data, or propose a correction, use that contact — or open an issue against
[CalCOFI/workflows](https://github.com/CalCOFI/workflows/issues) if you are
unsure who owns a dataset. Every dataset's provenance and open questions are
also tracked in that repo's `metadata/{provider}/{dataset}/questions.csv`.

## See also

- [The database](db.qmd) — table/column reference, including the
  `dataset` table's full column list.
- [Data Access](data-access.qmd) — how to query the release from R, Python or
  your browser.
- [Portals](portals.qmd) — where the same data is discoverable outside
  calcofi.io (EDI, NCEI, OBIS, ERDDAP).
