---
aliases:
  - helpers.html
---

# Access the data {#sec-access}

The integrated database is published as **parquet files on a public bucket**, and every way in —
R, Python, a SQL shell, a browser tab — reads those same bytes through the same catalog. There is
no server to be slow, no credentials to request and nothing to download whole: DuckDB reads only
the columns and row groups a query touches, over HTTPS. This chapter is the one guide to both
paths CalCOFI.io offers, the public release and the team's working store:

```{r}
#| label: setup
#| include: false
librarian::shelf(dplyr, gt, here, jsonlite, purrr, readr, tibble, quiet = TRUE)
`%||%` <- function(a, b) if (is.null(a)) b else a
snap    <- jsonlite::fromJSON(here("data/_snapshot.json"))
catalog <- jsonlite::fromJSON(here("data/release/catalog.json"), simplifyVector = FALSE)
ds      <- readr::read_csv(here("data/release/dataset.csv"), show_col_types = FALSE)
v       <- catalog$version
cov <- function(key) ds$coverage_temporal[ds$dataset_key == key] %||% "—"
tabs <- purrr::map_dfr(catalog$tables, \(t) tibble::tibble(
  table = t$name,
  tier  = dplyr::case_when(isTRUE(t$deprecated) ~ "deprecated", isTRUE(t$supplemental) ~ "supplemental", TRUE ~ "core"),
  rows  = as.numeric(t$rows %||% NA),
  objects = length(t$objects %||% list()),
  partitioned_by = {
    p <- unique(unlist(purrr::map(t$objects %||% list(), \(o) o$partition_by))); if (length(p)) paste(p, collapse = ", ") else ""
  },
  twin = any(purrr::map_lgl(t$objects %||% list(), \(o) is.null(o$partition_by))) && isTRUE(t$partitioned)))
```

| | the public release | the working store |
|---|---|---|
| what | every dataset, integrated, frozen per version, cited by DOI | the CTD team's multi-user PostgreSQL: originals verbatim, a flag ledger beside them |
| who | anyone | the CTD team and the data team |
| how | DuckDB over HTTPS: this page | an SSH tunnel: [Server access](server-access.qmd) |
| when it changes | at each release, never in place | continuously; accepted flags reach the next release |

## Where the data lives

A release is a versioned folder of **sidecars** — the record of what it is — and its tables are
parquet **objects** the release's `catalog.json` points at:

```
gs://calcofi-db/ducklake/
├── releases/
│   ├── latest.txt              # the promoted version
│   ├── versions.json           # every version: date, tables, rows, DOI; consolidated / retired
│   ├── RELEASES.md             # what changed between versions and why, newest first
│   └── {version}/
│       ├── catalog.json        # every table, its rows and its objects[] — where the bytes live
│       ├── metadata.json       # table and column descriptions, units, datasets, measurement types
│       ├── relationships.json  # primary and foreign keys        ┐ Keys and integrity
│       ├── integrity.json      # the keys, measured               ┘
│       ├── datasets.json       # one record per dataset — what every portal reads
│       ├── eml/, stac/, coverage.json, …
│       └── RELEASE_NOTES.md
└── tables/                     # the objects, content-addressed and shared between versions
    ├── {table}/{content_hash}/{table}.parquet
    └── {table}/{col}={value}/{content_hash}/data_0.parquet
```

The public HTTPS root is `https://storage.googleapis.com/calcofi-db/`; every `path` in a catalog is
relative to it (`https://storage.calcofi.io/calcofi-db/` serves the same bucket and redirects a
legacy per-release URL to its object where one exists). The promoted version is
[`releases/latest.txt`](https://storage.googleapis.com/calcofi-db/ducklake/releases/latest.txt);
what a version is, which are kept and how to pin one are in [Releases](releases.qmd).

**Resolve a table through the catalog; never build a path by hand.** Each table's `objects[]`
lists one parquet object for a single-file table and one per partition for a partitioned one,
each with its bucket-relative `path`, `bytes`, `sha256` and `since`. The packages' resolvers —
`calcofi4r::cc_release_sources()`, `calcofi4py.release_sources()` — turn a catalog entry into
URLs for any version, and the browser example below does it in six lines of JavaScript. A typed
`releases/{version}/parquet/…` path works only while that version's copy is kept, and the objects
it points at may be shared, moved or thinned; the catalog is the contract.

## What you are reading

Two tables hold the observations, and one view unites them:

- **`obs_bio`** — every biological occurrence headline: a taxon (`taxon_key`), its life stage, the
  measurement (`measurement_type`, `value`), the gear and effort of its own sampling event
  (`tow_type`, `std_haul_factor`, `prop_sorted`, `volume_sampled_m3`) and the two standardized
  densities (`density_per_10m2`, `density_per_1000m3`). One object; one taxon question is one
  file.
- **`obs_env`** — every environmental headline: one measured scalar per row, **one object per
  `measurement_type`**, so one variable is one small fetch — what a browser needs.
- **`obs`** — the two under one name, `realm` = `bio` or `env` and `value` as `measurement_value`,
  as a **view** the catalog carries (`views.obs`). `cc_get_db()` in R and Python and the query site
  create it, so `FROM obs` keeps working; its own objects are deprecated and drop in the next
  release.

Every observation points at its **sampling event** in `sample` (`sample_key`; `parent_sample_key`
walks net → tow → site and bottle → cast) and carries its provenance and place —
`dataset_key`, `cruise_key`, `grid_key`, position, time, `hex_id` — so a rollup needs no join.
Event-level effort such as `std_haul_factor` is in `sample_measurement`; sub-occurrence detail
(length bins, stages) in `obs_attribute`; the references are `cruise`, `ship`, `grid`, `taxon`,
`dataset`, `measurement_type`, `spatial`. [The database](db.qmd) describes each; [Keys and
integrity](keys.qmd) says how they join.

```{r}
#| label: tbl-tables
#| tbl-cap: !expr paste0("The tables of ", v, " as the catalog lays them out. A partitioned table reads as an explicit list of objects (with `hive_partitioning = true` to recover the partition column); a twin is one whole-table object a browser reads instead of the list.")
tabs |>
  dplyr::arrange(tier, table) |>
  dplyr::mutate(twin = ifelse(twin, "yes", "")) |>
  dplyr::select(table, tier, rows, objects, `partitioned by` = partitioned_by, twin) |>
  gt() |>
  fmt_integer(columns = c(rows, objects)) |>
  sub_missing(missing_text = "") |>
  tab_style(style = cell_text(font = "monospace", size = "0.85em"), locations = cells_body(columns = c(table, `partitioned by`))) |>
  tab_options(table.width = pct(100), data_row.padding = px(3), table.font.size = "0.88em")
```

A **supplemental** table (`obs_ctd_full`, the full-resolution CTD scans; `obs_mets_full`; the
`sample_root` join twin) is hosted and catalogued but left out of `cc_get_db()`'s default set —
`obs_ctd_full` alone is ~270 M rows. Ask for it (`supplemental = TRUE`) and filter early.

## From R

[`calcofi4r`](https://calcofi.io/calcofi4r/) registers every release table as a view over the
remote parquet — resolving each through the catalog, handling `httpfs`, the partitioned tables,
the `obs` view and a content-addressed local cache, so a table unchanged between releases is
fetched once:

```r
# remotes::install_github("calcofi/calcofi4r")
library(calcofi4r)
library(dplyr)

con <- cc_get_db()                       # the promoted release, every table as a view
con <- cc_get_db("v2026.08.25")          # pinned; a retired version errors, naming the replacement
con <- cc_get_db(supplemental = TRUE)    # + obs_ctd_full, obs_mets_full, sample_root
DBI::dbListTables(con)

tbl(con, "taxon") |> filter(scientific_name == "Sardinops sagax")          # lazy dbplyr
cc_query("SELECT dataset_key, count(*) AS n FROM obs_env GROUP BY 1 ORDER BY 2 DESC")

cc_list_versions()                       # versions.json
cc_release_notes()                       # this release's notes
cc_describe_table("sample")              # columns, units, descriptions from metadata.json
```

To hand a table's URLs to DuckDB (or anything else) yourself, resolve them through the catalog:

```r
cat_ <- cc_catalog("latest")                     # or a pinned "vYYYY.MM.DD"
src  <- cc_release_sources(cat_, "obs_env")      # list(urls, hive, canonical, hashes, local_paths, single_file)
src$urls                                         # one https URL per partition
cc_read_parquet_sql(src)                         # read_parquet([...], hive_partitioning = true) — paste into any SQL
```

## From Python

[`calcofi4py`](https://calcofi.io/calcofi4py/) mirrors the R package, verb for verb:

```python
# pip install "calcofi4py @ git+https://github.com/CalCOFI/calcofi4py"
import calcofi4py as cc

con = cc.cc_get_db()                              # the promoted release; cc_get_db("v2026.08.25") to pin
df  = con.sql("SELECT * FROM taxon WHERE common_name ILIKE '%sardine%'").df()
cc.cc_query("SELECT count(*) FROM obs_bio").fetchone()

catalog = cc.cc_catalog("latest")
src     = cc.release_sources(catalog, "obs_env")  # {"urls", "hive", "canonical", "hashes", "local_paths", "single_file"}
cc.read_parquet_sql(src)                          # the read_parquet(...) fragment for any DuckDB
```

## From SQL, anywhere

Any DuckDB — the command-line shell, a notebook, another language's binding — reads the release
after `INSTALL httpfs; LOAD httpfs;` (and `spatial` for `ST_*` functions). Take the table's
objects from the catalog and hand them to `read_parquet()`; a partitioned table is an explicit
list with `hive_partitioning = true`, and that runs in every DuckDB, the browser included:

```sql
INSTALL httpfs; LOAD httpfs;

-- one object: a single-file table (the path is catalog.json ▸ tables[name = 'taxon'] ▸ objects[0].path)
SELECT taxon_key, scientific_name, common_name, worms_id, rank
FROM read_parquet('https://storage.googleapis.com/calcofi-db/ducklake/tables/taxon/{content_hash}/taxon.parquet')
WHERE scientific_name = 'Sardinops sagax';

-- many objects: a partitioned table, one URL per partition, in catalog order
SELECT measurement_type, count(*) AS n
FROM read_parquet([
    'https://storage.googleapis.com/calcofi-db/ducklake/tables/obs_env/measurement_type=temperature/{hash}/data_0.parquet',
    'https://storage.googleapis.com/calcofi-db/ducklake/tables/obs_env/measurement_type=salinity/{hash}/data_0.parquet'
  ], hive_partitioning = true)
GROUP BY 1;
```

Because the partition column is in the path, a filter on it lets DuckDB **prune** whole objects:
a one-variable query over `obs_env` opens one file. Releases before v2026.09 have no `objects[]`;
their partitioned tables are directories that plain HTTPS cannot glob, so they are read through an
anonymous s3-style glob (`SET s3_endpoint = 'storage.googleapis.com'; SET s3_url_style = 'path';`
then `read_parquet('s3://calcofi-db/…/obs/**/*.parquet', hive_partitioning = true)`) — the
resolvers return that form for those versions automatically.

## From your browser

DuckDB compiles to WebAssembly, so the same engine runs in a browser tab with nothing installed.
Three ready-made ways in:

- the [**Explorer**](https://calcofi.io/explore/) — six lenses over the whole database, every view a
  URL, every download with its SQL ([Explore](explore.qmd));
- [**calcofi.io/db-query**](https://calcofi.io/db-query/) — a library of queries to start from,
  including the three biology-to-environment matches below, and a free SQL shell;
- [**shell.duckdb.org**](https://shell.duckdb.org) — DuckDB's own shell, for a one-off query.

To embed DuckDB in your own page, the boilerplate is short: load the bundle, instantiate, resolve
a table through the catalog:

```html
<script type="module">
  import * as duckdb from "https://cdn.jsdelivr.net/npm/@duckdb/duckdb-wasm@1.29.0/+esm";
  const bundle = await duckdb.selectBundle(duckdb.getJsDelivrBundles());
  const db     = new duckdb.AsyncDuckDB(new duckdb.ConsoleLogger(), await duckdb.createWorker(bundle.mainWorker));
  await db.instantiate(bundle.mainModule, bundle.pthreadWorker);
  const conn = await db.connect();
  await conn.query("INSTALL httpfs; LOAD httpfs;");

  const root    = "https://storage.googleapis.com/calcofi-db";
  const version = (await (await fetch(`${root}/ducklake/releases/latest.txt`)).text()).trim();
  const catalog = await (await fetch(`${root}/ducklake/releases/${version}/catalog.json`)).json();

  // a table's objects → a read_parquet() fragment. A partitioned table is a list (hive_partitioning);
  // if it also publishes a whole-table twin (the object WITHOUT partition_by), a browser reads the
  // twin instead of the list — never both, which would count every row twice.
  const readParquet = (name) => {
    const t    = catalog.tables.find(t => t.name === name);
    const twin = t.partitioned ? t.objects.find(o => !o.partition_by) : null;
    if (twin) return `read_parquet('${root}/${twin.path}')`;
    const urls = t.objects.map(o => `'${root}/${o.path}'`).join(", ");
    return `read_parquet([${urls}]${t.partitioned ? ", hive_partitioning = true" : ""})`;
  };

  const result = await conn.query(`
    SELECT scientific_name, common_name, worms_id FROM ${readParquet("taxon")}
    WHERE common_name ILIKE '%sardine%'`);
  console.log(result.toArray());
</script>
```

## Apply the quality flags

`measurement_qual` carries each dataset's **own** flag vocabulary, uninterpreted — bottle
(6 = OK but from CTD, 8 = suspect, 9 = missing), CTD cast files (1/2 = use the primary/secondary
sensor, 8 = questionable, 9 = bad/missing), DIC WOCE (2 = good, 3 = questionable, 4 = bad,
9 = missing); the registry is
[`metadata/measurement_qual.csv`](https://github.com/CalCOFI/workflows/blob/main/metadata/measurement_qual.csv).
A flagged value is still a row. `obs_bio` and `obs_env` carry the verdict as a column, **`qual_ok`**,
and the packages expose the predicate it was computed with, so a filter reads the same in every
language:

```sql
WHERE qual_ok                          -- obs_bio, obs_env: the materialized verdict
```
```r
calcofi4r::cc_qual_ok_sql("o")         # the predicate, for obs / obs_ctd_full / sample_measurement
```
```python
calcofi4py.qual_ok_sql("o")
```

The predicate is NULL-safe — an unflagged row is kept — and every CalCOFI app applies it. A 2.18 ml/L
oxygen spike at 1,144 m that had been flagged suspect since 1955 reached a plot in 2026 because no
consumer did; that is why the column exists.

## Matching biology to environment

Relating a net tow's catch to the water it was towed through is the question the database exists
to answer, and `calcofi4r` answers it in one call: three wrappers match **biological** observations
to **environmental** ones in time and space, on the fly, against the release.

| function | biological side |
|---|---|
| `cc_match_ichthyo_by_name()` | ichthyoplankton, filtered by scientific name |
| `cc_match_ichthyo_by_taxon()` | ichthyoplankton, a WoRMS taxon **and all its descendants** (a recursive walk of `taxon.parent_taxon_key`) |
| `cc_match_zooplankton_biomass()` | net-tow displacement-volume biomass (`totalplankton` / `smallplankton`) |

```r
d <- cc_match_ichthyo_by_name(
  "Sardinops sagax", env_var = "temperature", life_stage = "larva",
  date_min = "2018-01-01", date_max = "2018-03-31",
  relax_matching = TRUE)                 # 5 km / 72 h instead of the default 2 km / 6 h

cat(attr(d, "sql"))                      # the exact, portable SQL that produced it
str(attr(d, "query_meta"))               # release version, parameters, source URLs
```

Each returns one row per biological observation with the matched environmental value, and
attaches the **exact SQL** that produced it: paste `attr(d, "sql")` into the DuckDB shell, Python
or the query site and the same rows come back. `return_sql = TRUE` gives the SQL without running
it. The wrappers are thin shells over `cc_match_bio_env(bio, env, …)`, which takes a biological
and an environmental `SELECT` and performs the match — call it when the wrappers' filters are not
yours. Its knobs: `max_dist_km` and `max_time_hr` (2 km / 6 h; `relax_matching = TRUE` widens to
5 km / 72 h), `join_method` (`"nearest_time"`, `"nearest_dist"` or `"average"` over the window),
`life_stage`, `date_min`/`date_max`, `depth_m_min`/`depth_m_max`, `version`.

::: {.callout-note}
The example asks for **2018** because the environmental and biological series end at different
times in the release: the bottle series covers `r cov("calcofi_bottle")`, the ichthyoplankton series
`r cov("swfsc_ichthyo")` (measured, not typed — [The database](db.qmd)). Pick a period both cover.
:::

## Reproducible

Every query is plain SQL against immutable, versioned, public parquet, so a CalCOFI result is
reproducible by anyone who has the version and the SQL. Pin a **consolidated** version
([Releases](releases.qmd)) — its bytes are kept indefinitely — and keep the SQL: what the
matching helpers attach as `attr(d, "sql")`, what the Explorer copies beside every download, what
the query site shows in its SQL panel and what a hand-written query is are one and the same thing,
in R, Python, the shell and the browser.

::: {.callout-tip collapse="true"}
## The retired API, endpoint by endpoint

The Plumber API at `api.calcofi.io` served pre-built match tables from a PostgreSQL database that
no longer exists. Every endpoint has a replacement that reads the release directly:

| retired endpoint | now |
|---|---|
| `/variables` | `cc_list_measurement_types()` |
| `/timeseries`, `/cruise_lines`, `/cruise_line_profile`, `/raster` | SQL over `obs_env` and `sample`, or the Explorer's sections and contours |
| `/cruises` | `cc_read_cruise()` |
| `zooplankton_biomass` | `cc_match_zooplankton_biomass()` |
| `itis_ichthyodata` | `cc_match_ichthyo_by_taxon()` — WoRMS, recursive |
| `ichthyodata` | `cc_match_ichthyo_by_name()` |
| `relax_matching`, `cruiseymd_min/max`, `stage` | `relax_matching`, `date_min/max`, `life_stage` |

[Appendix A](api.qmd) keeps the old endpoint reference.
:::
