---
aliases:
  - apps.html
  - reports.html
  - maps.html
---

# Explore {#sec-explore}

Everything on this page runs on the same frozen release, so what you see in one product is what
you get from another. Start with the Explorer; the rest of the page is where to go when you want
one dataset, one table, or one query.

## The Explorer

[**calcofi.io/explore**](https://calcofi.io/explore/) is one web application for looking at the
whole integrated database, for one **organism** (a taxon from the net tows and censuses) or one
**ocean variable** (from the bottle, CTD, carbonate and weather series) at a time, through six
*lenses*:

- **Stations** — every CalCOFI station as a dot, coloured by the summary of what you chose over
  the years, season and depths you set.
- **Hexagons** — the same summary pooled into hexagons of a resolution you pick, so the whole
  survey area reads at once.
- **Contours** — an interpolated surface over the stations, computed in the browser, shown with
  its own error.
- **Cruises** — one cruise at a time: its track, its stations, what it measured.
- **Regions** — summaries by boundary: sanctuaries, protected areas, wind areas, the CalCOFI
  regions.
- **Sections** — a vertical slice along a line, station by depth, as values or as anomalies
  against the one climatology every product subtracts (see [The database](db.qmd)).

Four things make it different from an app with a server behind it:

- **Nothing runs anywhere but your browser.** The SQL executes in DuckDB inside the page, against
  the release's parquet files on a public bucket — the same bytes `calcofi4r` and `calcofi4py`
  read. There is no login, no rate limit and nothing to install.
- **The URL is the whole view.** Lens, variable, stage, years, season, depth, theme: share the
  address and the other person sees your view.
- **Every download is reproducible.** A map exports as PNG with its title; a table as CSV with a
  `dataset_key` column; and the SQL that produced it can be copied for R, Python or a DuckDB
  shell.
- **Attribution is built in.** *Cite this data* copies the release citation plus one per dataset
  in the view (see [Cite this data](cite.qmd)); a *Sources* line names every dataset a view pools;
  the feedback button sends the view, annotated if you like, to the team.

`?theme=dark` or `?theme=light` sets the theme, `?tour=off` skips the welcome tour, and the
[README](https://github.com/CalCOFI/explore#readme) documents every parameter.

## One dataset

Every dataset in the database — and every one CalCOFI holds but has not yet ingested — has a page
at [**calcofi.io/datasets**](https://calcofi.io/datasets/): what it is, who made it, its years and
area, how to cite it, and every place it can be reached (the Explorer, the packages, ERDDAP, its
archive of record). The page is generated from the same record the portals read, so it is never a
copy of anything. [Metadata & the ingest loop](metadata.qmd) says where its words come from.

## One table, one query

- [**calcofi.io/db-schema**](https://calcofi.io/db-schema/) browses the schema of any release:
  tables, columns with units and descriptions, the entity-relationship diagram, the datasets and
  the measurement-type registry. [The database](db.qmd) and [Keys and integrity](keys.qmd) are the
  narrative it lacks.
- [**calcofi.io/db-query**](https://calcofi.io/db-query/) runs SQL against the release in your
  browser — a library of queries to start from, and a shell. [Access the data](data-access.qmd)
  shows the same queries from R and Python.

## The other applications

```{r}
#| label: setup
#| include: false
librarian::shelf(dplyr, gt, here, purrr, yaml, quiet = TRUE)
`%||%` <- function(a, b) if (is.null(a)) b else a
prod  <- yaml::read_yaml(here("data/products.yml"))
cards <- prod$products %||% prod
card_row <- function(p) tibble::tibble(
  key         = p$key,
  title       = p$title %||% p$key,
  section     = p$section %||% NA_character_,
  group       = p$group %||% NA_character_,
  live_url    = p$live_url %||% NA_character_,
  description = p$description %||% "",
  status      = p$status %||% "live",
  superseded_by = p$superseded_by %||% NA_character_,
  credits     = if (is.null(p$credits)) NA_character_ else paste(unlist(p$credits), collapse = ", "),
  added       = p$added %||% NA_character_)
cards_df <- purrr::map_dfr(cards, card_row)
title_of <- function(k) { t <- cards_df$title[match(k, cards_df$key)]; ifelse(is.na(t), k, t) }
status_text <- function(s, by) dplyr::case_when(
  s == "superseded" & !is.na(by) ~ paste0("superseded by ", title_of(by)),
  s == "superseded" ~ "superseded",
  s == "archived"   ~ "archived",
  s == "interim"    ~ "interim",
  TRUE              ~ "live")
apps_tbl <- function(d) {
  d |>
    dplyr::mutate(Status = status_text(status, superseded_by),
                  Product = live_url, name = title) |>
    dplyr::select(Product, name, Status, description) |>
    (\(tbl) if (knitr::is_html_output())
      tbl |> gt() |> fmt_url(columns = "Product", label = from_column("name")) |> cols_hide("name")
    else
      tbl |> dplyr::select(-Product) |> dplyr::rename(Product = name) |> gt())() |>
    cols_label(description = "") |>
    cols_align("left") |>
    tab_options(table.width = pct(100), data_row.padding = px(5), table.font.size = "0.92em")
}
```

The Explorer superseded the server-side applications that each showed one of its lenses; they
keep serving, and their cards on [calcofi.io](https://calcofi.io) say what replaced them. The table
reads the same cards.

```{r}
#| label: tbl-apps
#| tbl-cap: The applications on calcofi.io besides the Explorer, from the landing page's product cards.
cards_df |>
  dplyr::filter(section == "explore", key != "explore") |>
  dplyr::arrange(status_text(status, superseded_by), title) |>
  apps_tbl()
```

## Student contributions

Student teams at UCSB, UCLA and UCSD have built portals, dashboards and stories on the database;
their repositories are archived and their sites keep serving.

```{r}
#| label: tbl-students
#| tbl-cap: Student contributions, from the landing page's product cards.
cards_df |>
  dplyr::filter(section == "students") |>
  dplyr::arrange(dplyr::desc(added), title) |>
  dplyr::mutate(description = ifelse(is.na(credits), description, paste0(description, " — ", credits))) |>
  apps_tbl()
```

## Reports built on the data

The [Channel Islands National Marine Sanctuary condition report](https://noaa-onms.github.io/cinms)
reads CalCOFI through `calcofi4r` — its [forage assemblage
page](https://noaa-onms.github.io/cinms/modals/forage-assemblage.html) is a worked example of a
report that regenerates from the release. The [status chapter](status.qmd) records what each
contract period delivered.
