# Naming conventions {#sec-naming}

These are the database's conventions: what a table, a column, a key, a measured quantity, a life
stage, a gear code, a provider or a dataset is called, and why. They are **generated from the
registries** in [CalCOFI/workflows](https://github.com/CalCOFI/workflows) `metadata/` — the same
files the ingest notebooks conform to and the release is checked against — so this chapter is the
convention as enforced, not as remembered. The companion for anyone *supplying* data is
[Providing data to CalCOFI](provide.qmd), which says which of these a provider needs to care about.

```{r}
#| label: setup
#| include: false
librarian::shelf(dplyr, gt, here, jsonlite, purrr, readr, stringr, tibble, tidyr, quiet = TRUE)
reg <- function(f) readr::read_csv(here("data/registry", f), show_col_types = FALSE, na = "")
snap  <- jsonlite::fromJSON(here("data/_snapshot.json"))
fld   <- reg("field_dictionary.csv")
mt    <- reg("measurement_type.csv")
cat_  <- reg("category.csv")
ls_   <- reg("life_stage.csv")
gear  <- reg("gear.csv")
prov  <- reg("provider.csv")
lic   <- reg("license.csv")
ds    <- readr::read_csv(here("data/release/dataset.csv"), show_col_types = FALSE)
mono  <- function(g, cols) tab_style(g, style = cell_text(font = "monospace", size = "0.85em"), locations = cells_body(columns = {{ cols }}))
compact <- function(g) tab_options(g, table.width = pct(100), data_row.padding = px(3), table.font.size = "0.86em")
```

## The rules

- **Tables are singular and lower case**: `cruise`, `sample`, `taxon` (not `cruises`).
- **Columns are `snake_case`**, lower case with underscores, so nothing ever needs quoting in
  SQL; a unit is a suffix where one applies (`depth_m`, `volume_sampled_m3`, `temperature_degc`).
- **The suffix of an identifier says what it holds**: `*_key` a string natural key, `*_id` an
  integer, `*_uuid` the provider's own identifier, `*_seq` a sequence. The four keys that carry the
  integration — `cruise_key`, `sample_key`, `site_key` / `grid_key`, `taxon_key` — and what each
  promises are in [Keys and integrity](keys.qmd).
- **Every measured value is a row, not a column**: `measurement_type`, `value` (`measurement_value`
  on the deprecated `obs`), `measurement_qual`, with the vocabulary of types registered once
  (below), so a new quantity is a registry row, never a schema change.
- **Text is UTF-8**; a geometry column is `geom`, tagged `EPSG:4326`; times are UTC (the
  dictionary names an event's start and end, `datetime_start_utc` / `datetime_end_utc`; the released
  `sample` carries the start as `datetime`); positions are decimal degrees, signed; depth is metres,
  positive down.

## Fields

The field dictionary is **prescriptive**: a new dataset's columns are mapped onto these names at
ingest, and the aliases record what each source called them. `dwc_term` is the Darwin Core term a
field publishes as, filled only where one term means exactly this (`r sum(nzchar(fld$dwc_term))`
of `r nrow(fld)` today).

```{r}
#| label: tbl-fields
#| tbl-cap: The field dictionary (`metadata/field_dictionary.csv`).
fld |>
  dplyr::mutate(dwc = ifelse(nzchar(dwc_term), basename(dwc_term), "")) |>
  dplyr::arrange(category, fld_new) |>
  dplyr::select(field = fld_new, type = type_new, units, category, description = fld_description, aliases, dwc) |>
  gt(groupname_col = "category") |>
  sub_missing(missing_text = "") |>
  mono(c(field, type, units, aliases, dwc)) |>
  compact()
```

## Measured quantities

`measurement_type.csv` is the vocabulary of everything measured, filed under the twelve
categories every dataset and type belongs to. A type carries its units, the physical bounds a
value must lie within (`valid_min` / `valid_max` — enforced per dataset at ingest, and a release
fails on a value outside them), the depth range over which it is defined, how a *derived* type
was produced, and — where one concept says exactly this — the NERC P01 concept and P06 unit a
Darwin Core export emits. `is_canonical` marks the headline types the apps offer first.

```{r}
#| label: tbl-categories
#| tbl-cap: The twelve categories (`metadata/category.csv`).
cat_ |>
  dplyr::arrange(order) |>
  dplyr::mutate(types = purrr::map_int(category, \(k) sum(mt$category == k, na.rm = TRUE)),
                datasets = purrr::map_int(category, \(k) sum(ds$category == k, na.rm = TRUE))) |>
  dplyr::select(category, realm, description, `measurement types` = types, datasets) |>
  gt() |> mono(category) |> compact()
```

```{r}
#| label: tbl-measurement-types
#| tbl-cap: !expr paste0("Every measurement type (", nrow(mt), "), by category. Bounds are physical limits, not observed ranges; a blank NERC column means no concept states exactly this, never that it was not looked at.")
mt |>
  dplyr::mutate(
    canonical = ifelse(coalesce(as.logical(is_canonical), FALSE), "yes", ""),
    bounds = dplyr::case_when(
      !is.na(valid_min) & !is.na(valid_max) ~ paste0(valid_min, " – ", valid_max),
      !is.na(valid_min) ~ paste0("≥ ", valid_min),
      !is.na(valid_max) ~ paste0("≤ ", valid_max),
      TRUE ~ ""),
    p01 = ifelse(nzchar(coalesce(nerc_p01, "")), basename(nerc_p01), "")) |>
  dplyr::arrange(category, dplyr::desc(canonical), measurement_type) |>
  dplyr::select(category, `measurement type` = measurement_type, canonical, units, bounds, description, `P01` = p01) |>
  gt(groupname_col = "category") |>
  sub_missing(missing_text = "") |>
  mono(c(`measurement type`, units, bounds, P01)) |>
  compact()
```

## Life stages and gear

Two more registries carry the values that appear on observations and events, each with the
Darwin Core label and the NERC concept where one is exact. Two life-stage values are recorded as
*not* life stages (euphausiid `damaged`, ichthyoplankton `invert`) and are never published as one.

```{r}
#| label: tbl-life-stages
#| tbl-cap: Life stages (`metadata/life_stage.csv`).
ls_ |>
  dplyr::mutate(s11 = ifelse(nzchar(coalesce(nerc_s11, "")), basename(nerc_s11), "")) |>
  dplyr::select(`life stage` = life_stage, `Darwin Core` = dwc_lifeStage, `NERC S11` = s11, parent = life_stage_parent, datasets, note) |>
  gt() |> sub_missing(missing_text = "") |> mono(c(`life stage`, `NERC S11`, datasets)) |> compact()
```

```{r}
#| label: tbl-gear
#| tbl-cap: Gear (`metadata/gear.csv`) — every `sample.tow_type` code.
gear |>
  dplyr::mutate(l22 = ifelse(nzchar(coalesce(nerc_l22, "")), basename(nerc_l22), "")) |>
  dplyr::select(code = tow_type, gear = gear_name, `sampling protocol` = dwc_samplingProtocol, `NERC L22` = l22, datasets) |>
  gt() |> sub_missing(missing_text = "") |> mono(c(code, `NERC L22`, datasets)) |> compact()
```

## Providers, licences and dataset names

**A provider is the organization that curates — holds and can license — the data**, never the
portal that serves it and never a collection or lab inside it. CalCOFI program data is `calcofi`
even when served from NCEI or EDI; the Pelagic Invertebrate Collection is a *dataset* of provider
`sio`. Every provider an ingest names must be registered, or the build stops.

```{r}
#| label: tbl-providers
#| tbl-cap: Providers (`metadata/provider.csv`).
prov |>
  dplyr::filter(coalesce(status, "active") != "deprecated") |>
  dplyr::mutate(datasets = purrr::map_int(provider, \(k) sum(ds$provider == k))) |>
  dplyr::select(provider, name = provider_name, url, status, datasets) |>
  (\(tbl) if (knitr::is_html_output()) tbl |> gt() |> fmt_url(columns = url, label = "↗")
          else tbl |> dplyr::select(-url) |> gt())() |>
  sub_missing(missing_text = "") |> mono(provider) |> compact()
```

**A dataset's key is `{provider}_{dataset}`**, both slugs lower case with hyphens inside a slug
(`cce-lter_zooscan`, `cdfw_dungeness-crab`). The key names everything about the dataset: its
ingest notebook, its folder of metadata, its row in the provider's Sheet, its record in the
release and its page at `calcofi.io/datasets/{key}/`. The human names — `dataset_name` and the
short form the apps show — are declared once, in the ingest notebook's `calcofi:` block, and every
surface reads them from there. This is the nomenclature the data management plan's Task 13 asks to
be agreed; the table is the current state of it.

```{r}
#| label: tbl-datasets
#| tbl-cap: !expr paste0("The datasets of ", snap$version, " — key, names, provider and category.")
ds |>
  dplyr::arrange(provider, dataset) |>
  dplyr::select(dataset_key, `name` = dataset_name, `short name` = dataset_name_short, provider, category, license) |>
  gt() |> sub_missing(missing_text = "") |> mono(c(dataset_key, provider, category, license)) |> compact()
```

```{r}
#| label: tbl-licenses
#| tbl-cap: Licence ids a dataset may declare (`metadata/license.csv`); `custom` needs a `license_url`.
lic |>
  dplyr::select(license, name, url, status, notes) |>
  (\(tbl) if (knitr::is_html_output()) tbl |> gt() |> fmt_url(columns = url, label = "↗")
          else tbl |> dplyr::select(-url) |> gt())() |>
  sub_missing(missing_text = "") |> mono(license) |> compact()
```

## Legacy names, mapped

Every dataset arrived with its own names, and every one of them is recorded: the dictionary's
`aliases` column lists what each standard field was called in the sources, and each ingest keeps a
crosswalk, `metadata/{provider}/{dataset}/flds_redefine.csv`, from every source column to its
standard name and type. Nothing is renamed in a source file; the mapping is applied at ingest and
kept so the provider's own names remain answerable.

```{r}
#| label: setup-crosswalk
#| include: false
xw_files <- list.files(here("data/registry/flds_redefine"), pattern = "\\.csv$", full.names = TRUE)
xw <- purrr::map_dfr(xw_files, \(f) {
  x <- readr::read_csv(f, show_col_types = FALSE, na = "", col_types = readr::cols(.default = "c"))
  need <- c("tbl_new", "fld_old", "fld_new", "type_old", "type_new", "units")
  for (k in setdiff(need, names(x))) x[[k]] <- NA_character_
  x |>
    dplyr::mutate(dataset_key = sub("\\.csv$", "", basename(f))) |>
    dplyr::select(dataset_key, table = tbl_new, `source name` = fld_old, `standard name` = fld_new,
                  `source type` = type_old, type = type_new, units)
})
xw_summary <- xw |>
  dplyr::group_by(dataset_key) |>
  dplyr::summarise(fields = dplyr::n(),
                   renamed = sum(!is.na(`source name`) & !is.na(`standard name`) & `source name` != `standard name`),
                   .groups = "drop")
```

```{r}
#| label: tbl-crosswalk-summary
#| tbl-cap: Each dataset's crosswalk — how many source fields it maps, and how many change name on the way in. The full crosswalk per dataset follows (HTML), and each is a CSV in CalCOFI/workflows.
xw_summary |>
  dplyr::mutate(crosswalk = paste0("https://github.com/CalCOFI/workflows/blob/main/metadata/",
                                   sub("_", "/", dataset_key, fixed = TRUE), "/flds_redefine.csv")) |>
  dplyr::arrange(dataset_key) |>
  (\(tbl) if (knitr::is_html_output())
    tbl |> gt() |> fmt_url(columns = crosswalk, label = "flds_redefine.csv")
  else
    tbl |> dplyr::select(-crosswalk) |> gt())() |>
  mono(dataset_key) |> compact()
```

::: {.content-visible when-format="html"}

```{r}
#| label: crosswalks
#| echo: false
#| results: asis
# one tab per dataset; only the fields whose name changes are listed, since an unchanged
# name needs no map. HTML only — a 600-row appendix would double the pdf.
cat("::: {.panel-tabset}\n\n")
for (k in sort(unique(xw$dataset_key))) {
  d <- xw |>
    dplyr::filter(dataset_key == k, !is.na(`source name`), `source name` != coalesce(`standard name`, "")) |>
    dplyr::arrange(table, `standard name`) |>
    dplyr::select(-dataset_key)
  cat("## ", k, "\n\n", sep = "")
  if (nrow(d) == 0) { cat("*Every source field keeps its name.*\n\n"); next }
  print(d |> gt(groupname_col = "table") |> sub_missing(missing_text = "") |>
          mono(c(`source name`, `standard name`, `source type`, type, units)) |> compact() |> as_raw_html())
  cat("\n\n")
}
cat(":::\n")
```

:::

## Proposing a change

A convention changes the way the data does: through a pull request to `metadata/` in
[CalCOFI/workflows](https://github.com/CalCOFI/workflows) — a row in the dictionary (authored in
`libs/build_field_dictionary.R`, which regenerates the CSV), a measurement type registered with
`calcofi4db::register_measurement_types()`, a life stage or gear row — or, where the question is
the provider's to answer, a row in that dataset's `questions.csv` and its Sheet
([Metadata & the ingest loop](metadata.qmd)). The next release carries whatever changed, and this
page follows.
