# Portals

## Overview

CalCOFI data is available through various portals, each serving different purposes and user
needs. Since 2026-09, every dataset — integrated into the database or still a holding — also has
one calcofi.io page ([`/datasets/{dataset_key}/`](https://calcofi.io/datasets/)) joining every
endpoint it can be reached through: the release Parquet, a CF netCDF, ERDDAP, and the archive or
aggregator portals below, each with its own publication status. That page, not this chapter, is
now the record — the tables here are generated from the same registries the page reads, so the
two can never disagree.

## Data Flow

While it would be ideal for CalCOFI data to be available through a single portal, each portal has
its strengths and limitations. The following diagram illustrates one possible realization of data
flow between CalCOFI data and the portals: from raw data to the integrated database to the dataset
catalog to the portals and meta-portals.

```{mermaid}
%%| label: fig-portals-flow
%%| fig-cap: "Flow of data from raw to database to the dataset catalog to portals and meta-portals."
%%| file: diagrams/portals_flow.mmd
%%| fig-width: 6
```

In practice, CalCOFI is a partnership with various contributing members, so the authoritative
dataset might flow differently, such as from EDI to the database to the other portals. The other
portals, such as OBIS or ERDDAP, serve different audiences or purposes. The dataset catalog is
what now feeds the meta-portals directly — ODIS crawls its sitemap and JSON-LD, Google Dataset
Search its JSON-LD, and data.gov its DCAT-US `data.json` — rather than each meta-portal having to
know about every downstream portal individually.

## Portal capabilities

While some portals serve as data repositories, others provide advanced data access and
visualization tools, and a few are meta-portals that index the others rather than holding data
themselves. The table below is generated from
[`CalCOFI/workflows`' `metadata/portal.csv`](https://github.com/CalCOFI/workflows/blob/main/metadata/portal.csv)
— the same registry `datasets.json`'s `registrations[]` and this book both read — so a portal
added there (CalOOS and the IOOS Catalog joined in 2026-09) shows up here without anyone
hand-editing a table.

```{r}
#| label: tbl-portals
#| tbl-cap: Portal capabilities, from `metadata/portal.csv` (CalCOFI/workflows).
#| warning: false

librarian::shelf(
  dplyr, glue, gt, here, purrr, readr, scales, tidyr)

portals_csv <- here("data/portal.csv")

# capability columns are "none" | "partial" | "full"; harvests_from_us and kind are
# plain text (a mechanism id, or "none"), shown without the symbol treatment
cap_cols <- c("full_archive", "versioning", "doi_issued", "query_by_xyt",
             "query_by_taxa", "multiple_formats", "api_access")
cap_symbols <- c("none" = "✖", "partial" = "▲", "full" = "✔")
cap_colors  <- c("none" = "#FFCCCC", "partial" = "#FFF2CC", "full" = "#CCFFCC")

# display label for each capability column, in the order they'll appear
cap_labels <- c("Full Archive", "Versioning", "DOI Issued", "Query by xyt",
                "Query by taxa", "Multiple formats", "API Access")

portals_tbl <- read_csv(portals_csv, show_col_types = FALSE) |>
  mutate(Portal = glue::glue("[{name}]({url})")) |>
  select(Portal, kind, all_of(cap_cols), harvests_from_us)
names(portals_tbl) <- c("Portal", "Kind", cap_labels, "Harvests from us")

portals_tbl |>
  gt(rowname_col = "Portal") |>
  fmt_markdown(columns = "Portal") |>
  data_color(
    columns = all_of(cap_labels),
    fn = scales::col_factor(palette = as.character(cap_colors), levels = names(cap_colors))) |>
  cols_align(align = "center", columns = everything()) |>
  text_transform(
    locations = cells_body(columns = all_of(cap_labels)),
    fn = function(x) unname(cap_symbols[x])) |>
  tab_source_note(source_note = md(
    "**Capability legend**: ✔ = full, ▲ = partial, ✖ = none. **Harvests from
    us** names the mechanism CalCOFI feeds that portal through (an ERDDAP WAF, a sitemap of
    JSON-LD, an IPT upload, …), or `none` for a discovery-only aggregator with no CalCOFI-fed
    channel yet.")) |>
  tab_options(table.width = pct(100), data_row.padding = px(5))
```

### EDI
**Environmental Data Initiative**

-   Complete dataset archives using DataOne software and EML metadata
-   DOIs issued for all datasets ensuring citability
-   Full archive allowing for any data file types
-   Basic spatial and temporal filtering through web interface
-   Download in original formats with metadata
-   Access through DataOne API
-   Links:
    -   [EDIrepository.org](https://edirepository.org)
    -   CalCOFI datasets: [EDI query "CalCOFI"](http://portal.edirepository.org:80/nis/simpleSearch?defType=edismax&q=CalCOFI&fq=-scope:ecotrends&fq=-scope:lter-landsat*&fl=id,packageid,title,author,organization,pubdate,coordinates&debug=false)


**Publishing there is manual, in four steps** (`publish_to-edi.qmd`, generic per dataset; the
package is built from the release's own `eml/{dataset_key}.xml` plus CSV entities exported from
the core — for the bottle database three tables totalling 2.1 GB plus the shared vocabulary
table). (1) An EDI account for the open `edi` scope, held by Erin, with Ben granted editor access
through the EML `access` element — created once at dashboard.edirepository.org. (2) Set
`EDI_USER`/`EDI_PASS` (or an EDI API key as `EDI_KEY`) wherever the pipeline runs and render the
notebook: PASTA **evaluate** runs automatically against **staging** and the report lands beside the
package; fix what it names and re-render. (3) Once evaluate is clean, render with
`CALCOFI_PUBLISH_EDI=true` to **create** (or, on a later release, **update**) the package against
**production** — this mints a DOI and a `scope.identifier.revision`, recorded in
`metadata/edi_packages.csv`. (4) Add the DOI back as an `archive` row in `metadata/distribution.csv`
(`portal: edi`) and as the dataset's `doi` in its `metadata/{provider}/{dataset}/dataset_meta.yml`,
so the record and the citation carry it from the next release on. CCE-LTER's `knb-lter-cce`
packages stay theirs and are never republished.

### NCEI
**National Centers for Environmental Information**

-   Long-term archival of oceanographic data
-   DOIs issued for dataset submissions
-   Standardized metadata using ISO 19115-2
-   Basic search interface with geographic and temporal filtering
-   Data preserved in original submission formats
-   Access through NCEI API services
-   Links:
    -   [NCEI Ocean Archive](https://www.ncei.noaa.gov/products/ocean-archive)
    -   CalCOFI datasets: [NCEI search "CalCOFI"](https://www.ncei.noaa.gov/access/search/index?q=CalCOFI)

### OBIS
**Ocean Biodiversity Information System**

-   Specialized in marine biodiversity data
-   Standardized using DarwinCore fields
-   Extended measurements supported via [extendedMeasurementOrFact](https://manual.obis.org/format_emof.html)
-   Powerful filtering by space, time, and taxonomic parameters
-   Multiple download formats (CSV, JSON, Darwin Core Archive)
-   Full REST API access
-   Published through the OBIS-USA Integrated Publishing Toolkit (`ipt-obis.gbif.us`), one
    resource per dataset whose taxa resolve to WoRMS
-   Links:
    -   [OBIS.org](https://obis.org)
    -   `swfsc_ichthyo` on OBIS: [*CalCOFI Fish Larvae & Egg Tows*](https://obis.org/dataset/0e223f55-c826-4513-ae9a-b04cbf2e189c)
        (77,188 events, 463,655 occurrences, 610,816 eMoF; featured by OBIS 2026-04-15) — the
        [dataset page](https://calcofi.io/datasets/swfsc_ichthyo/) carries the rest as they
        publish


**Uploading is manual, and deliberately so.** `publish_to-obis.qmd` writes
`data/darwincore/{dataset_key}/{dataset_key}_{version}.zip` and a `{dataset_key}_manifest.json`;
nothing in the pipeline talks to a portal. CalCOFI has no IPT of its own — the resources live on the
**OBIS-USA IPT**, `ipt-obis.gbif.us`, under Ben's login. Before a first upload, the OBIS records that
already cover the same source are listed (the notebook queries `api.obis.org/v3/dataset` and prints
the curated rows from `metadata/distribution.csv`) and resolved **with their owners**: a provider's
own record is never duplicated, and a historical CalCOFI record is retired or cross-referenced only
with its owner's agreement. Then: open (or create) the resource → upload the zip's files as the
source → map the Event core plus the Occurrence and ExtendedMeasurementOrFact extensions exactly as
`meta.xml` describes them → publish a new version. Afterwards, record the OBIS dataset id and the
IPT resource id as `distribution.csv` rows (`portal = obis` / `ipt`, `kind = archive`) and stamp
`uploaded_utc` + `uploaded_hash` into the manifest. That last step is what makes a dataset page
honest: the record reports `built, not uploaded` / `published (vX)` / `stale — data changed in vY` by
comparing the uploaded hash with the archive's own, so it says when an upload is due. Re-upload only
when the `content_hash` changed — it is deterministic over the archive's CSV bytes.

### ERDDAP
**Environmental Research Division Data Access Program**

-   Tabular and gridded data server
-   Advanced subsetting by space, time, and parameters
-   Multiple output formats (CSV, JSON, NetCDF, etc.)
-   RESTful API with direct data access
-   Built-in data visualization tools
-   No persistent identifiers but stable URLs
-   Links:
    -   [ERDDAP](https://upwell.pfeg.noaa.gov/erddap/index.html)
    -   [erddap.calcofi.io](https://erddap.calcofi.io/erddap/index.html) — CalCOFI's own
        ERDDAP over the release, every dataset's `infoUrl` pointing back at its
        [dataset page](https://calcofi.io/datasets/)
    -   [ERDDAP, OceanView - CalCOFI seabirds](https://oceanview.pfeg.noaa.gov/erddap/search/index.html?page=1&itemsPerPage=1000&searchFor=seabird+CalCOFI)
    -   [ERDDAP, CoastWatch - CalCOFI oceanographic](https://coastwatch.pfeg.noaa.gov/erddap/search/index.html?page=1&itemsPerPage=1000&searchFor=CalCOFI)

### Zenodo

-   One DOI per integrated-database release, via the GitHub integration on
    [CalCOFI/workflows](https://github.com/CalCOFI/workflows) — a concept DOI covers every
    version, a version DOI each one
-   Archives the pipeline itself at the release tag, plus `catalog.json` / `metadata.json`
-   Link: [zenodo.org](https://zenodo.org)

### CalOOS

**The SCCOOS + CeNCOOS regional data portal**

-   Registering erddap.calcofi.io gives every existing CoastWatch-fed module a calcofi.io
    distribution beside it
-   What data.ioos.us (the national IOOS Catalog) harvests for this region — likely the only
    IOOS-level registration CalCOFI needs
-   Link: [data.caloos.org](https://data.caloos.org)

## Metadata

The [Ecological Metadata Language (EML)](https://eml.ecoinformatics.org/) is the one metadata
document CalCOFI's publishers share. Since 2026-09 the release writes one EML 2.2 document per
dataset — `eml/{dataset_key}.xml`, built by `calcofi4db::build_eml()` directly from
`datasets.json` and each dataset's descriptive sidecar, checked by `check_eml()` before it ships —
rather than each publish notebook typing its own sampling methods, study extent and rights by
hand. The Darwin Core Archive OBIS publishing uses and an EDI data package both start from the
same document, so a provider correction to a dataset's abstract or creators reaches every
publisher at once.

```{mermaid}
%%| label: fig-portals-metadata
%%| fig-cap: "Metadata in the form of ecological metadata language (EML) is used to describe the dataset in a consistent manner that can be ingested by the portals."
%%| file: diagrams/portals_metadata.mmd
%%| fig-width: 6
```

The EML specification provides detailed structure for describing datasets, including:

-   [Dataset identification and citation](cite.qmd)
-   Geographic and temporal coverage
-   Variable definitions and units
-   Methods and protocols
-   Quality control procedures
-   Access and usage rights

This standardized metadata enables automated data transformation and ingestion into various portal
systems while preserving the original data context and quality information.

## Meta-Portals

### Google Dataset Search

The JSON-LD metadata on every calcofi.io dataset page gets indexed by
[Google Dataset Search](https://datasetsearch.research.google.com/search?query=calcofi) through
schema.org metadata. This ensures that CalCOFI data is discoverable through Google search and
other search engines.

### ODIS
**Ocean Data Information System**

ODIS uses the same technology as Google Dataset Search (schema.org, JSON-LD), but focuses on
ocean data. ODISCat record 3318 points at calcofi.io's generated
[`datasets/sitemap.xml`](https://calcofi.io/datasets/sitemap.xml), so a dataset's calcofi.io page
is what [ODIS.org](https://odis.org) crawls — the record's `sameAs` then links each external
portal record from there.

### data.gov (DCAT-US)

calcofi.io publishes a [`data.json`](https://calcofi.io/data.json) (Project Open Data / DCAT-US
1.1), generated from `datasets.json` the same way the sitemap is — one entry per public dataset,
with its licence, distributions and contact point. This is what a CKAN instance (data.gov, a
university's, CDFW's) harvests with no work on our side.

This federated approach ensures that CalCOFI data remains:

-   Discoverable through multiple channels
-   Properly cited and attributed
-   Integrated with global ocean data systems

## The dataset catalog, and who registers where

`datasets.json` (`calcofi4db::build_dataset_catalog()`, calcofi4db ≥ 4.1.0) is the record this
whole chapter is a rendering of: one entry per dataset, its `registrations[]` naming every portal
above with a status of `published`, `planned` (usually linking the tracking issue) or `n/a`. The
table below reads it **live**, straight from the promoted release, so it can never drift from what
`calcofi.io/datasets/{dataset_key}/`'s own Access table shows.

```{r}
#| label: tbl-registrations
#| tbl-cap: Per-dataset portal registrations, read live from the promoted release's `datasets.json`.
#| warning: false
#| results: asis

librarian::shelf(dplyr, glue, gt, jsonlite, purrr, scales, tidyr)

release_base <- "https://storage.googleapis.com/calcofi-db/ducklake/releases"
latest_version <- tryCatch(
  trimws(readLines(file.path(release_base, "latest.txt"), warn = FALSE)[1]),
  error = function(e) NA_character_)

catalog_record <- if (!is.na(latest_version)) tryCatch(
  jsonlite::fromJSON(file.path(release_base, latest_version, "datasets.json"), simplifyVector = TRUE),
  error = function(e) NULL) else NULL

reg_status_symbols <- c("n/a" = "✖", "planned" = "▲", "published" = "✔")
reg_status_colors  <- c("n/a" = "#FFCCCC", "planned" = "#FFF2CC", "published" = "#CCFFCC")

if (is.null(catalog_record) || is.null(catalog_record$datasets)) {
  version_label <- if (is.na(latest_version)) "unresolved" else latest_version
  cat(glue::glue(
    "> The promoted release ({version_label}) carries no `datasets.json` yet ",
    "(the dataset catalog ships from calcofi4db >= 4.1.0). This table renders automatically once ",
    "one does — see `https://storage.googleapis.com/calcofi-db/ducklake/releases/latest.txt`.\n"))
} else {
  ds <- catalog_record$datasets
  reg_long <- purrr::map2_dfr(ds$dataset_key, ds$registrations, function(key, regs) {
    if (is.null(regs) || !nrow(regs)) return(NULL)
    tibble::tibble(dataset_key = key, portal = regs$portal, status = regs$status)
  })
  reg_wide <- reg_long |>
    tidyr::pivot_wider(names_from = portal, values_from = status, values_fill = "n/a") |>
    dplyr::left_join(dplyr::select(ds, dataset_key, dataset_name_short), by = "dataset_key") |>
    dplyr::mutate(Dataset = glue::glue(
      "[{dataset_name_short}](https://calcofi.io/datasets/{dataset_key}/)")) |>
    dplyr::select(Dataset, dplyr::everything(), -dataset_key, -dataset_name_short) |>
    dplyr::arrange(Dataset)
  portal_cols <- setdiff(names(reg_wide), "Dataset")

  reg_wide |>
    gt(rowname_col = "Dataset") |>
    fmt_markdown(columns = "Dataset") |>
    data_color(
      columns = all_of(portal_cols),
      fn = scales::col_factor(palette = as.character(reg_status_colors), levels = names(reg_status_colors))) |>
    cols_align(align = "center", columns = everything()) |>
    text_transform(
      locations = cells_body(columns = all_of(portal_cols)),
      fn = function(x) unname(reg_status_symbols[x])) |>
    tab_source_note(source_note = md(glue::glue(
      "**Status legend**: ✔ = published, ▲ = planned, ✖ = n/a. Read from ",
      "`{latest_version}`'s `datasets.json`; the same rows back each dataset's own page."))) |>
    tab_options(table.width = pct(100), data_row.padding = px(5))
}
```

## Getting the data through calcofi.io

The portals above are where CalCOFI data is discoverable *outside* calcofi.io; calcofi.io itself
is a fourth way in, and it is where the record lives:

-   **[The dataset catalog](https://calcofi.io/datasets/)** — one page per dataset (and per
    holding not yet ingested), with its coverage, every distribution above, its citation and
    provenance, and machine surfaces (`{key}.json`, `{key}.jsonld`, the sitemap and `data.json`
    described above).
-   **[Data Access](data-access.qmd)** — query the release Parquet directly from R, Python or a
    browser with DuckDB; no credentials, no API server.
-   **Libraries** — `calcofi4r::cc_datasets()` / `calcofi4py.cc_datasets()` read the same
    `datasets.json` this page's tables do, into a data frame; `cc_get_db()` / `cc_query()` for the
    data itself.
-   **[Cite This Data](cite.qmd)** — `cc_cite()` in `calcofi4r` / `calcofi4py` for exactly the
    datasets a query touched, checked citations, licences and DOIs.
-   **[Apps](apps.qmd)** — interactive exploration (the Explorer, Station and Hexagon Explorers,
    CTD Transects) without writing a query at all; each app's Sources panel links back to the same
    dataset pages.
