# Products, brand and uptime {#sec-products}

Every CalCOFI.io product is a card on [calcofi.io](https://calcofi.io), and the card is the
record: its name, its section, where it lives, where its source is, its uptime history and its
usage. This page lists them from that record and states the three small contracts that keep a
product findable and recognisably CalCOFI's. For the data team and anyone building a new one.

```{r}
#| label: setup
#| include: false
librarian::shelf(dplyr, gt, here, jsonlite, purrr, tibble, yaml, quiet = TRUE)
`%||%` <- function(a, b) if (is.null(a)) b else a
prod  <- yaml::read_yaml(here("data/products.yml"))
secs  <- purrr::map_dfr(prod$sections %||% list(), \(s) tibble::tibble(section = s$key %||% s$id %||% NA_character_, title = s$title %||% NA_character_))
cards <- purrr::map_dfr(prod$products %||% list(), \(p) tibble::tibble(
  key = p$key, title = p$title %||% p$key, section = p$section %||% NA_character_,
  live_url = p$live_url %||% NA_character_, source_url = p$source_url %||% NA_character_,
  uptime = p$uptime %||% NA_character_, usage = p$usage %||% NA_character_,
  status = p$status %||% "live", superseded_by = p$superseded_by %||% NA_character_,
  added = p$added %||% NA_character_,
  tech = if (is.null(p$tech)) "" else paste(unlist(p$tech), collapse = ", ")))
title_of <- function(k) { t <- cards$title[match(k, cards$key)]; ifelse(is.na(t), k, t) }
cards <- cards |>
  dplyr::mutate(
    section_title = secs$title[match(section, secs$section)] %||% section,
    status_text = dplyr::case_when(
      status == "superseded" & !is.na(superseded_by) ~ paste0("superseded by ", title_of(superseded_by)),
      TRUE ~ status),
    uptime_url = ifelse(is.na(uptime), NA, paste0("https://status.calcofi.io/history/", uptime)),
    usage_url  = ifelse(is.na(usage),  NA, paste0("https://calcofi.io/analytics/", usage, "/")))
```

## Every product

@tbl-products is every card on the landing page, by section, with where the product lives, where
its source is and how it is doing.

```{r}
#| label: tbl-products
#| tbl-cap: The products on calcofi.io, by section, from the landing page's cards (`_data/products.yml` in CalCOFI/CalCOFI.github.io). Usage per product is at [calcofi.io/analytics](https://calcofi.io/analytics/).
d <- cards |>
  dplyr::arrange(factor(section, levels = unique(secs$section)), status_text, title) |>
  dplyr::select(section = section_title, key, title, live_url, source_url, uptime_url, status = status_text, tech)
if (knitr::is_html_output()) {
  d |>
    gt(groupname_col = "section") |>
    fmt_url(columns = live_url,   label = from_column("title")) |>
    fmt_url(columns = source_url, label = "source") |>
    fmt_url(columns = uptime_url, label = "uptime") |>
    cols_hide(title) |>
    cols_label(live_url = "product", source_url = "", uptime_url = "") |>
    sub_missing(missing_text = "") |>
    tab_style(style = cell_text(font = "monospace", size = "0.85em"), locations = cells_body(columns = c(key, tech))) |>
    tab_options(table.width = pct(100), data_row.padding = px(3), table.font.size = "0.88em")
} else {
  d |>
    dplyr::select(section, key, title, status, tech) |>
    gt(groupname_col = "section") |>
    sub_missing(missing_text = "") |>
    tab_options(table.width = pct(100), data_row.padding = px(3), table.font.size = "0.88em")
}
```

## The ship's log

calcofi.io/news merges four kinds of entry into one history, newest first — releases, datasets
entering the release, new products, and hand-written notes — generated by `_plugins/news.rb` in
`CalCOFI/CalCOFI.github.io` and republished as the Atom feed `/feed.xml`; the header's News link
wears a dot while any entry is under 30 days old. Three of the four are read off records this book
already reads; only the fourth — a feature, a change to the site, a paper — is typed by hand, in
the site's `_data/news.yml`. A hand-written row carrying `version:` or `dataset_key:` replaces the
generated entry of that key, for when the generated title is not the story.

@tbl-ship-log is what each kind is: its source, who is responsible for it appearing, when it shows
up, and where it sends a reader — with how many of each this snapshot carries.

```{r}
#| label: tbl-ship-log
#| tbl-cap: The ship's log's four kinds of entry, and how many of each this snapshot carries.
rel_versions <- jsonlite::fromJSON(here("data/release/versions.json"), simplifyVector = TRUE)$versions
rel_datasets <- jsonlite::fromJSON(here("data/release/datasets.json"), simplifyVector = FALSE)$datasets
news_rows    <- yaml::read_yaml(here("data/news.yml"))

n_release <- sum(!is.na(rel_versions$release_date))
since_v   <- purrr::map_chr(rel_datasets, \(d) d[["since_version"]] %||% NA_character_)
n_dataset <- sum(since_v %in% rel_versions$version)
n_app     <- sum(!is.na(cards$added))
news_type <- purrr::map_chr(news_rows, \(r) r[["type"]] %||% NA_character_)
n_hand    <- sum(news_type %in% c("data", "site", "paper"))

log_tbl <- tibble::tibble(
  type              = c("release", "dataset", "app", "data · site · paper"),
  count             = c(n_release, n_dataset, n_app, n_hand),
  source            = c(
    "versions.json + the first ## heading of the version's section in RELEASES.md",
    "datasets[].since_version in the record",
    "products.yml added:",
    "_data/news.yml"),
  `who writes it`   = c(
    "the release runner, in RELEASES.md # Unreleased before the cut",
    "nobody — the release that first carries the dataset",
    "whoever adds the card",
    "whoever has the story: a feature, a change to the site, a paper"),
  `when it appears` = c(
    "minutes after latest.txt is promoted (test_release.qmd dispatches refresh.yml), or the Monday cron",
    "with that release",
    "on push",
    "on push"),
  `where it links`  = c(
    "RELEASES.html#v{version}",
    "the dataset page",
    "the product",
    "the row's url"))

log_tbl |>
  gt() |>
  fmt_integer(columns = count) |>
  tab_style(style = cell_text(font = "monospace", size = "0.85em"), locations = cells_body(columns = c(source, `where it links`))) |>
  tab_options(table.width = pct(100), data_row.padding = px(3), table.font.size = "0.88em")
```

## The three-slug contract

A product is one thing in three places, and it is the same string in all of them:

- its `key` in the landing page's `products.yml` (the card);
- its `slug` in `CalCOFI/uptime`'s monitor list (the status page at
  `status.calcofi.io/history/<slug>`);
- its `slug` in `CalCOFI/analytics`' registry (the usage report at `calcofi.io/analytics/<slug>/`).

When the names diverged, none of the three was findable from the others. Cards are named for the
thing itself — the repository or application name used everywhere else — except student
contributions, which keep their human titles. The card's `status` is its *lifecycle* (`interim`,
`superseded`, `archived`), not its uptime; a superseded card names what replaced it, and the
[Explore](explore.qmd) chapter reads exactly that.

## The brand contract

Every product wears one theme, one header and one favicon, and the rule is checked weekly rather
than assumed. The contract is [calcofi.io/brand/v2](https://calcofi.io/brand/v2/) — the SIO look,
light by default, UCSD's palette, Source Sans 3 and Teko, the horizontal lockup — and in one
breath:

- `?theme=dark|light` on any URL sets the theme, remembered across products only when a person
  chose it; a fresh visit is light.
- The header is one bar: the lockup (the mark and the wordmark "CalCOFI.io") at the far left links
  to calcofi.io, the product's own name beside it ("Explorer", "Query", never "CalCOFI Query") links
  to its own root, the theme toggle sits at the right. Where a framework owns the bar
  (Quarto, pkgdown, MkDocs, Shiny), the lockup goes in its brand slot and its toggle is bridged.
- An application declares its scale (`<meta name="cc-scale" content="app">`); a page declares
  nothing and gets the reading rhythm. This book is a page.
- `?tour=off` suppresses any guided tour, so `live_url?theme=<t>&tour=off` is a deterministic
  screenshot; every card carries a light and a dark shot taken that way.

`calcofi4r::cc_brand_head()` and `cc_brand_header()` give a Shiny app the head block and the bar;
the Quarto sites include the same head file; `check_brand.py` in the landing repository probes
every product for the theme, the favicon, the home link, the toggle, the version and the default.

## Adding a product

1. Build it on the release: read through the catalog ([Access the data](data-access.qmd)), never a
   typed path; apply the quality predicate; cite through `cc_cite()`.
2. Wear the brand (above); honour `?theme=` and `?tour=off`; take the two screenshots.
3. Add the card to `products.yml` with its section, its `datasets` (the keys it covers, or `all`)
   and, if it opens a dataset directly, the `dataset_url` template the dataset pages link through.
4. Add the same slug to the uptime monitors and the analytics registry, and tag the page for
   analytics.
5. Push; the landing page, the status page and the usage report rebuild themselves, and this table
   follows.
