# Releases {#sec-releases}

The integrated database is published as **frozen releases**: an immutable, versioned copy of every
table, with a catalog that says exactly which bytes it holds and a DOI that names it. Nothing a
consumer reads is ever edited in place. This chapter says what a release is, what ships with it,
which versions are kept, and what is checked before one is cut.

```{r}
#| label: setup
#| include: false
librarian::shelf(dplyr, gt, here, jsonlite, purrr, readr, stringr, tibble, yaml, quiet = TRUE)
`%||%` <- function(a, b) if (is.null(a)) b else a
snap     <- jsonlite::fromJSON(here("data/_snapshot.json"))
catalog  <- jsonlite::fromJSON(here("data/release/catalog.json"), simplifyVector = FALSE)
versions <- jsonlite::fromJSON(here("data/release/versions.json"), simplifyVector = TRUE)$versions
policy   <- yaml::read_yaml(here("data/registry/release_policy.yml"))
tests    <- tryCatch(jsonlite::fromJSON(here("data/release/test_results.json")), error = function(e) NULL)
rel_md   <- readLines(here("data/release/RELEASES.md"), warn = FALSE)
fmt_n    <- function(x) format(x, big.mark = ",", scientific = FALSE)
fmt_gb   <- function(b) sprintf("%.2f GB", b / 1e9)
tabs <- purrr::map_dfr(catalog$tables, \(t) tibble::tibble(
  table = t$name,
  rows = as.numeric(t$rows %||% NA),
  objects = length(t$objects %||% list()),
  partition_by = purrr::map_chr(list(t$objects %||% list()), \(o) {
    p <- unique(unlist(purrr::map(o, \(x) x$partition_by))); if (length(p)) paste(p, collapse = ",") else ""
  }),
  bytes = sum(purrr::map_dbl(t$objects %||% list(), \(o) as.numeric(o$bytes %||% 0))),
  since = {
    s <- unlist(purrr::map(t$objects %||% list(), \(o) o$since)); if (length(s)) max(s) else NA_character_
  },
  supplemental = isTRUE(t$supplemental), deprecated = isTRUE(t$deprecated)))
```

## A release in one line

The promoted release is **`r catalog$version`**, cut `r catalog$release_date`: `r length(catalog$tables)`
tables, `r fmt_n(catalog$total_rows)` rows, `r fmt_gb(catalog$total_size)` of parquet, DOI
[`r catalog$doi`](https://doi.org/`r catalog$doi`). A version is a date (`vYYYY.MM.DD`), and a
release is cut whenever the data or the schema change enough to matter — every few weeks, in
practice. `latest.txt` on the bucket names the promoted one, and it is promoted only after the
consumer-contract suite and the client packages' README examples (`calcofi4r`, `calcofi4py`)
have passed against the frozen files.

## What ships

Everything lives under one public prefix, `gs://calcofi-db/ducklake/releases/{version}/`, read
over HTTPS with no credentials:

- **`catalog.json`** — the structural manifest and the data-access contract: the version, its
  DOIs and citation, and for every table its rows, its `content_hash` and its `objects[]` — one
  parquet object for a single-file table, one per partition for a partitioned one — each with a
  bucket-relative `path`, `bytes`, `sha256` and `since`. Resolve a table through it
  (`calcofi4r::cc_release_sources()`, `calcofi4py.release_sources()`); never build a path by hand.
- **the parquet objects**, content-addressed under `gs://calcofi-db/ducklake/tables/{table}/…`:
  immutable and shared between versions, so `since` on an object is the first release that shipped
  those exact bytes — the per-table changelog. A release's own `parquet/` copy exists only for the
  promoted version, its predecessor and the consolidated milestones (below).
- **the sidecars**: `metadata.json` (every table and column, with units and descriptions),
  `relationships.json` (every key, [Keys and integrity](keys.qmd)), `integrity.json` (the keys
  measured), `datasets.json` (one record per dataset — the record every portal reads), `eml/`
  (one EML document per dataset), `coverage.json` and `coverage_stations.json` (what the Explorer
  paints first), `grid.geojson`, `spatial.geojson`, `spatial_layers.json`, a static `stac/`
  catalog, `erd.mmd`, `test_results.json`, and `RELEASE_NOTES.md`.

@tbl-objects is that catalog rendered: every table of the promoted release, how many objects
carry it, and the version its bytes last changed in.

```{r}
#| label: tbl-objects
#| tbl-cap: !expr paste0("The tables of ", catalog$version, " as the catalog describes them — objects, partitioning, size, and the release whose bytes each table last changed in.")
tabs |>
  dplyr::mutate(tier = dplyr::case_when(deprecated ~ "deprecated", supplemental ~ "supplemental", TRUE ~ "core")) |>
  dplyr::arrange(tier, table) |>
  dplyr::select(table, tier, rows, objects, `partitioned by` = partition_by, bytes, `last changed` = since) |>
  gt() |>
  fmt_integer(columns = c(rows, objects)) |>
  fmt_bytes(columns = bytes, standard = "decimal", decimals = 1) |>
  sub_missing(missing_text = "") |>
  tab_style(style = cell_text(font = "monospace", size = "0.85em"), locations = cells_body(columns = c(table, `partitioned by`))) |>
  tab_options(table.width = pct(100), data_row.padding = px(3), table.font.size = "0.88em")
```

## Versions, and which are kept

Every version keeps its sidecars forever, so any release can still be browsed. Its parquet copy is
kept for the promoted version, the one before it, and the **consolidated** milestones — versions a
paper, a docs example or an app pins; the rest is thinned, and a thinned version's `retired`
marker names the consolidated one to read instead (`metadata/release_policy.yml` in
`CalCOFI/workflows` decides; `scripts/thin_releases.R` applies it). The content-addressed store
means a kept object is stored once however many versions share it. @tbl-versions is every version,
newest first, with what is kept of each.

```{r}
#| label: tbl-versions
#| tbl-cap: Every release, newest first, from `versions.json`.
# `retired` is a nested data frame column when any version carries it (jsonlite simplifies it)
ret_to <- if (!is.null(versions$retired)) versions$retired$to else rep(NA_character_, nrow(versions))
vv <- tibble::tibble(
  version = versions$version, date = versions$release_date, tables = versions$tables,
  rows = versions$total_rows, `size (MB)` = versions$size_mb,
  DOI = if (!is.null(versions$doi)) versions$doi else NA_character_,
  kept = dplyr::case_when(
    dplyr::coalesce(versions$consolidated, FALSE) ~ "consolidated",
    !is.na(ret_to) ~ paste0("retired → ", ret_to),
    TRUE ~ "current")) |>
  dplyr::arrange(dplyr::desc(version))
consolidated_str <- paste0("`", unlist(policy$consolidated), "`", collapse = ", ")
vv |>
  gt() |>
  fmt_integer(columns = c(tables, rows)) |>
  fmt_number(columns = `size (MB)`, decimals = 0) |>
  sub_missing(missing_text = "") |>
  tab_options(table.width = pct(100), data_row.padding = px(3), table.font.size = "0.88em")
```

Consolidated today: `r consolidated_str`.

To pin a version rather than follow the promoted one:

```r
calcofi4r::cc_get_db(version = "v2026.08.25")          # R
```

```python
calcofi4py.cc_get_db(version="v2026.08.25")             # Python
```

```r
# any DuckDB, from any language: resolve a table's objects through the catalog and paste the
# read_parquet() fragment it hands back — never a typed releases/{v}/parquet path
calcofi4r::cc_read_parquet_sql(calcofi4r::cc_release_sources(calcofi4r::cc_catalog("v2026.08.25"), "sample"))
```

## Two DOIs

Each release is tagged on GitHub, and Zenodo mints a **version DOI** for it; one **concept DOI**
always resolves to the newest. Cite the version DOI in a paper — it names exactly the bytes you
used — and the concept DOI where a living pointer is wanted. Both are in `catalog.json`, in the
release notes' *How to cite*, and on [Cite this data](cite.qmd):

- this release: <https://doi.org/`r catalog$doi`>
- every release: <https://doi.org/`r catalog$concept_doi`>

## The changelog

`RELEASES.md` is the database's NEWS file: one section per release, newest first, saying what
changed and why, with a **Consumers:** line wherever a change could break a reader, and an
`# Unreleased` section collecting changes until the next cut. Each version's `RELEASE_NOTES.md`
is its section plus a generated appendix (tables and rows, datasets, the consumer-contract result,
package versions). A release with no section for itself is refused before the freeze — the notes
are written by the change that made them necessary, never reconstructed afterwards.

The rendered changelog is [RELEASES.html](https://storage.calcofi.io/calcofi-db/ducklake/releases/RELEASES.html),
with the raw markdown at [RELEASES.md](https://storage.googleapis.com/calcofi-db/ducklake/releases/RELEASES.md)
as a secondary link. Every version heading is anchored `#v{version}`. The first `##` heading of a
version's section becomes that release's headline on calcofi.io/news and in the Atom feed — so it
is written as the story of the release.

```{r}
#| label: releases-headings
#| echo: false
#| results: asis
heads <- rel_md[grepl("^# ", rel_md)]
heads <- heads[!grepl("release notes$", heads, ignore.case = TRUE)]
cat("The most recent sections of the [rendered changelog](https://storage.calcofi.io/calcofi-db/ducklake/releases/RELEASES.html) (raw markdown at [`RELEASES.md`](https://storage.googleapis.com/calcofi-db/ducklake/releases/RELEASES.md)):\n\n")
for (h in head(heads, 6)) cat("- ", sub("^# ", "", h), "\n", sep = "")
```

## What a release checks before it ships {#sec-release-gates}

`release_database.qmd` assembles every ingest's parquet shards in memory, builds the release-only
tables, and runs a series of gates before the freeze. Each is a function in `calcofi4db` with a
test of its own, and the notebook's rendered page shows every result. The ones that stop a release
outright:

{{< include _gates.qmd >}}

: The gates that stop a release outright. {#tbl-gates}

And the ones that report and ratchet — a number that may only ever go down between releases, so a
regression fails but a known backlog does not:

| gate | what it measures |
|---|---|
| `check_depth_bounds()` / `check_depth_vs_seafloor()` | no depth is NaN, negative or beyond 6,500 m (fails); each root sample's deepest depth against the GEBCO seafloor near its position (ratchet — the place is imprecise, not the measurement) |
| `check_measurement_bounds()` | every measurement type has declared physical bounds and no value lies outside them (out of range fails; undeclared types ratchet) |
| `check_taxon_ids()` | every taxon key is an accepted WoRMS or ITIS id, or on an explicit allowlist of dataset-local classes (fails) |
| `check_dataset_citation()` | every dataset's citation has a year and a locator, its licence is a registered id, its DOI resolves (fails unless a provider question is open on the field) |
| `check_dataset_catalog()` / `check_eml_catalog()` | the record and each EML document validate against their schemas (fails) |
| `observed_coverage()` | each dataset's temporal and spatial extent, measured from the data — never typed (report) |

: The gates that report and ratchet rather than stop the release. {#tbl-ratchets}

A failure in @tbl-gates stops the release; a number in @tbl-ratchets that went up since the last
release stops it too, and one that went down becomes the new ceiling.

```{r}
#| label: contract-result
#| echo: false
#| results: asis
if (!is.null(tests)) cat(sprintf(
  "For %s the consumer-contract suite ran %d queries: **%d passed**, %d failed, %d skipped (`test_results.json`).\n",
  tests$release_version, tests$n_pass + tests$n_fail + tests$n_skip, tests$n_pass, tests$n_fail, tests$n_skip))
```

A release that passes is uploaded, tested once more from the bucket by `test_release.qmd`, and
only then promoted; the read-only consumers are refreshed afterwards by
`scripts/deploy_consumers.sh` — the Explorer, ERDDAP and the Shiny apps on the CalCOFI server,
and the ones hosted on GitHub Actions, which the script dispatches: `db-viz-station`,
`ctd-transects` and **this book**, whose pre-render snapshots the promoted release's
`catalog.json`, `integrity.json` and sidecars and so describes the previous version until it
re-renders ([Architecture](architecture.qmd)). A **staging** run exercises the whole path under a separate prefix without touching
the promoted release or the shared object store.
